---
title: "plot clades gaia"
output: html_document
date: "2026-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

load libraries
```{r}
#library(gaia)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(stringr)
library(purrr)
library(scales)
library(gridExtra)
library(rsample)
library(broom)
library(tibble)

```

ggplot theme
```{r}
## ggplot theme

my_theme <- function() {
    theme_classic() +
    theme( 
        plot.title = element_text(color="black", size=12, face="bold"),
        axis.title.x = element_text(color="black", size=20, face="bold", vjust=2),
        axis.title.y = element_text(color="black", size=20, face="bold", vjust=2),
        axis.text=element_text(size=24),
        )
}

options(repr.plot.width=14, repr.plot.height=8)
```

```{r}
jscores <- read.csv("clade_scores_jaccard_002.csv")

## filter out the pm 0.01 x mr 1e-4 
jscores <- jscores %>% filter(!mr == 0.000100)

jscores <- jscores %>% mutate(rr = mr*pm) 


jscores$pm <- factor(jscores$pm,
                                      levels=c(0.00100, 0.07575, 0.15050, 0.22530, 0.30000),
                                      labels = c(".001", ".076", ".151", ".226", "0.3" ) )

jscores$mr <- factor(jscores$mr, 
                                      levels=c( 0.000010, 0.001258, 0.002505, 0.003753, 0.005000), 
                                      labels = c( "1e-5", "1.25e-3", "2.5e-3", "3.75e-3", "5e-3" ) )


```


```{r}
pms <- unique(jscores$pm)
mrs <- unique(jscores$mr)
```


# plot node_height by node_height_inf
```{r, fig.width = 12, fig.height = 8}

jscores %>% filter(gaia_target == 'inferred') %>% ggplot(aes(x = node_height_inf, y = node_height, color = seed)) + 
                        geom_point() + 
                        theme_classic() + 
                        facet_wrap(pm~mr) +
                        geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed", linewidth = 1) +
                        scale_x_log10() + scale_y_log10() + 
                        ylab("true") + xlab("inferred") + ggtitle("true vs. inferred clade height")
                      

```

# proportion of clades accurately reconstructed 
```{r, fig.width = 12, fig.height = 5}
#, fig.width = 12}

## proportion below or above jaccard .8 
matches <- jscores %>% filter(gaia_target == 'inferred')  %>% mutate(above_80 = case_when(jaccard < .8 ~ 0, 
                                                                                          jaccard >= .8 ~ 1))

matches %>% group_by(seed, pm, mr, mu) %>% 
            summarize(proportion_reconstructed = mean(match)) %>% 
            ggplot(aes(x = proportion_reconstructed)) + 
            geom_histogram(binwidth = 0.01) + 
            facet_wrap(~as.factor(mu), nrow = 1) +
            theme_classic() + 
            geom_vline(xintercept = 0.8, color = 'darkgrey', linetype = 'dashed') + 
            xlab('proportion reconstructed clades per run')

matches %>% 
            group_by(seed, pm, mr, mu) %>% 
            summarize(proportion_reconstructed = mean(jaccard)) %>% 
            ggplot(aes(x = proportion_reconstructed)) + 
            geom_histogram(binwidth = 0.01) + 
            facet_wrap(~as.factor(mu),  nrow = 1) +
            theme_classic() + 
            geom_vline(xintercept = 0.8, color = 'darkgrey', linetype = 'dashed') + 
            xlab('mean jaccard index per run')


```

# clade reconstruction across genomic positions 
```{r, fig.width = 12, fig.height = 8}

# bin_matches <- jscores %>% filter(gaia_target == 'inferred') %>% group_by(pm, mu, bin, position) %>% summarize(proportion_reconstructed = mean(match))
# 
# mus = unique(bin_matches$mu)
# 
# plot_list <- list() 
# 
# for (i in 1:length(pms)) {
#   
#   run <- bin_matches %>% ungroup() %>% filter((pm == pms[i]) & (mu == mus[i]))
#   #s <- sample(unique(run$seed), 1)
#   #run <- run %>% filter(seed == s)
# 
#   p <- run %>% 
#           ggplot(aes(y = proportion_reconstructed, x = position)) + 
#                         geom_point(color = seed) + 
#                         geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
#                         #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
#                         theme_classic() + ggtitle(paste0("p/m: ", pms[i], ", mu: ", mus[i]))
#   
#   plot_list[[i]] <- p
# }
# 
# do.call(grid.arrange, c(plot_list, nrow = 2, ncol = 3  ))

```


```{r}

# number of pm/mr 
jscores %>% group_by(pm, mr, seed) %>% summarise(count = n()) %>% groupby(pm, mr) %>% summarise(count = n())

# length(list.files('../trees')) 1218 --> 608 total runs excluding test run 
```

## extract root height by finding max node_height per-bin 
later: crude scoring of runs according to root labeling accuracy 
```{r}
#roots <- jscores %>% group_by(bin, mr, mu, pm, seed, gaia_target) %>% summarise(root_height = max(node_height_sim, na.rm = TRUE, 
#                                                                                 mean_score = ....)) 

```


# plot accuracy x nodeheight across pm and mr 
```{r, fig.width = 12, fig.height = 8}

jscores %>% group_by(node_height, gaia_target, pm, mr) %>% summarise(score = mean(score)) %>% ungroup() %>% ggplot(aes(x=node_height, y=score, color=gaia_target)) +
                    geom_point(alpha = 0.5) +
                    facet_wrap(pm~mr) +
                    #facet_wrap(~mr) +
                    labs(x="node_depth", y="accuracy", color="gaia target") +
                    scale_x_log10() +
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) +
                    theme_classic() #+ theme(legend.position = "none")

```

# binomial logistic regression: accuracy~node_depth

### get predicted scores for each node height using logistic regression 
```{r}

run_summary_grouped <- function(grouped_data, group_key) {
  run <- grouped_data
  model <- glm(score ~ log10(node_height), family = "binomial", data = run)
  augmented_run <- broom::augment(model, run) %>% mutate(probability = plogis(.fitted))
  return(augmented_run) 
}

jresult_df <- jscores %>% filter(match == 1) %>% 
  group_by(pm, mr, seed, gaia_target) %>% 
  group_modify(~run_summary_grouped(.x, .y)) 

```

### plot decay curves for a few situations (individual runs)
```{r}


plot_list <- list() 
# 
# result_df <- jscores %>% filter(match == 1) %>% 
#   group_by(pm, mr, seed, gaia_target) %>% 
#   group_modify(~run_summary_grouped(.x, .y)) 

#par(mfrow = c(3, 2))
for (i in 1:length(pms)) {
  
  run <- jresult_df %>% ungroup() %>% filter((pm == pms[i]) & (mr == mrs[i]) & (gaia_target == 'simulated'))
  s <- sample(unique(run$seed), 1)
  run <- run %>% filter(seed == s)
  # model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
  # run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))
  
  p <- run %>% 
          ggplot(aes(x=node_height, y=score)) + 
                      geom_point(alpha = 0.15) + 
                      geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                      geom_line(aes(y = probability), linewidth = 1) + 
                      #facet_wrap(~mr) + 
                      labs(x="node_depth", y="accuracy") + 
                      scale_x_log10() + 
                      geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                      #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                      theme_classic() + ggtitle(paste0("p/m: ", pms[i], ", mr: ", mrs[i]))
  
  plot_list[[i]] <- p
}

do.call(grid.arrange, c(plot_list, nrow = 2, ncol = 3  ))

```

# simulated - 1 per param situation (just geom_line fits)
```{r, fig.width = 15, fig.height = 10}

plot_list <- list() 
count <- 1

for (i in 1:length(pms)) {
  for (j in 1:length(mrs)) {
  
  run <- jresult_df %>% ungroup() %>% filter((pm == pms[i]) & (mr == mrs[j]) & (gaia_target == 'simulated'))
  s <- sample(unique(run$seed), 1)
  run <- run %>% filter(seed == s)

  p <- run %>% 
          ggplot(aes(x=node_height, y=score)) + 
                      geom_point(alpha = 0.15) + 
                      geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                      geom_line(aes(y = probability), linewidth = 1) + 
                      #facet_wrap(~mr) + 
                      labs(x="node_depth", y="accuracy") + 
                      scale_x_log10() + 
                      geom_hline(yintercept = .80,  linetype = "dashed", color = "grey") + #geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                      #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                      theme_classic() + ggtitle(paste0("p/m: ", pms[i], ", mr: ", mrs[j]))
  
  plot_list[[count]] <- p
  count <- count + 1
  }
}
do.call(grid.arrange, c(plot_list, nrow = 5, ncol = 5 ))

```



# overlay with inferred run, fit 
```{r, fig.width = 15, fig.height = 10}

plot_list <- list() 
count <- 1

for (i in 1:length(pms)) {
  for (j in 1:length(mrs)) {
  
  run <- jresult_df %>% ungroup() %>% filter((pm == pms[i]) & (mr == mrs[j]) & (match == 1))
  s <- sample(unique(run$seed), 1)
  run <- run %>% filter(seed == s)

  p <- run %>% 
          ggplot(aes(x=node_height, y=score, color = gaia_target)) + 
                      geom_point(alpha = 0.15, aes(color = gaia_target)) + 
                      geom_smooth(method = "glm", method.args = list(family = "binomial"), aes(fill = gaia_target)) +
                      geom_line(aes(y = probability), linewidth = 1) + 
                      #facet_wrap(~mr) + 
                      labs(x="node_depth", y="accuracy") + 
                      scale_x_log10() + 
                      geom_hline(yintercept = .80,  linetype = "dashed", color = "grey") +# geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                      scale_color_manual(values = c("simulated" = "black", "inferred" = "#00A08A")) + 
                      scale_fill_manual(values = c("simulated" = "#606060", "inferred" = "#00A08A")) + 
                      theme_classic() + ggtitle(paste0("p/m: ", pms[i], ", mr: ", mrs[j])) + theme(legend.position = 'none')
  
  plot_list[[count]] <- p
  count <- count + 1
  }
}
do.call(grid.arrange, c(plot_list, nrow = 5, ncol = 5 ))

```


## predict accuracy across grid of node heights. get upper bound, lower bound, node_height threshold for which score drops below 80%; 
```{r}
run_summary_grouped <- function(grouped_data, group_key) {
  # fit model 
  model <- glm(score ~ log10(node_height), family = "binomial", data = grouped_data)
  
  # grid of node_heights
  grid_heights <- seq(min(grouped_data$node_height), max(grouped_data$node_height) * 1.5, length.out = 5000)
  grid_df <- data.frame(node_height = grid_heights)
  
  # predict scores on grid 
  grid_preds <- predict(model, newdata = grid_df, type = "link", se.fit = TRUE)
  
  # calculate probabilities and confidence bounds for run
  grid_df <- grid_df %>%
    mutate(
      probability  = plogis(grid_preds$fit),
      lower_bound = plogis(grid_preds$fit - (1.96 * grid_preds$se.fit)),
      upper_bound = plogis(grid_preds$fit + (1.96 * grid_preds$se.fit))
    )
  
  # actual threshold where p crosses .80
  # model_fits <- data.frame(node_height = grouped_data$node_height, 
  #                          fit = predict(model, type = "response")) %>% 
  #                          arrange(node_height)
  # h_empirical <- model_fits$node_height[which(model_fits$fit <= 0.80)[1]]
  

  # find the first node height in the grid where each curve drops below 0.80
  h_prob  <- grid_df$node_height[which(grid_df$probability <= 0.80)[1]]
  h_lower <- grid_df$node_height[which(grid_df$lower_bound <= 0.80)[1]]
  h_upper <- grid_df$node_height[which(grid_df$upper_bound <= 0.80)[1]]
  
  h_prob50  <- grid_df$node_height[which(grid_df$probability <= 0.50)[1]]
  h_lower50 <- grid_df$node_height[which(grid_df$lower_bound <= 0.50)[1]]
  h_upper50 <- grid_df$node_height[which(grid_df$upper_bound <= 0.50)[1]]

  # predict on actual run data
  orig_preds <- predict(model, newdata = grouped_data, type = "link", se.fit = TRUE)
  
  # add the bounds from grid prediction to run df
  grouped_data %>%
    mutate(
      probability = plogis(orig_preds$fit),
      lower_bound = plogis(orig_preds$fit - (1.96 * orig_preds$se.fit)),
      upper_bound = plogis(orig_preds$fit + (1.96 * orig_preds$se.fit)),
      #empirical_node_height_at_80 = h_empirical,
      mean_node_height_at_80  = h_prob,
      lower_node_height_at_80 = h_lower,
      upper_node_height_at_80 = h_upper,
      mean_node_height_at_50  = h_prob50,
      lower_node_height_at_50 = h_lower50,
      upper_node_height_at_50 = h_upper50
    ) %>%
    arrange(node_height)
}

jscore_df4 <- jscores %>%
  filter(match == 1) %>%
  group_by(pm, mr, seed, gaia_target) %>%
  group_modify(~run_summary_grouped(.x, .y)) %>%
  ungroup()


jscore_df4$gaia_target <- factor(jscore_df4$gaia_target, levels = c('simulated', 'inferred'))

```

example node height threshold for individual runs (sim)
```{r, fig.width = 15, fig.height = 10}

plot_list <- list() 
count <- 1


for (i in 1:length(pms)) {
  for (j in 1:length(mrs)) {
  
  run <- jscore_df4 %>% ungroup() %>% filter((pm == pms[i]) & (mr == mrs[j]) & (gaia_target == 'simulated'))
  s <- sample(unique(run$seed), 1)
  run <- run %>% filter(seed == s)

  p <- run %>% 
          ggplot(aes(x=node_height, y=score)) + 
                      geom_point(alpha = 0.1) + 
           
                      geom_line(aes(y = probability), linewidth = 1) + 
      
                      labs(x="node_depth", y="accuracy") + 
                      scale_x_log10() + 
                      geom_hline(yintercept = .80, color = "grey") + 
               
                      geom_vline(xintercept = unique(run$mean_node_height_at_80), color = 'lightblue') + #predicted
                      
                      geom_rect(xmin = unique(run$lower_node_height_at_80), # shading the confidence intervals) 
                                xmax = unique(run$upper_node_height_at_80), 
                                #alpha = 0.8, 
                                ymin = 0, ymax = 1, 
                                fill = 'blue') + 

                      theme_classic() + ggtitle(paste0("p/m: ", pms[i], ", mr: ", mrs[j]))
  
  plot_list[[count]] <- p
  count <- count + 1
  }
}

#plot_list
do.call(grid.arrange, c(plot_list, nrow = 5, ncol = 5 ))
```

# plot node depth threshold (0.8) with increasing mr
```{r, fig.width = 12, fig.height = 3}

## boxplots where each point describes one run and each box describes an mr 

jscore_df4 %>% 
          ggplot(aes(x=factor(mr), y=mean_node_height_at_80, fill = gaia_target)) +
          #geom_jitter(width = 0.1, alpha = 0.1, size = 0.1) + 
          geom_boxplot(alpha = 0.4) + 
          facet_wrap(~pm, nrow = 1) + 
          labs(x="migration strength", y="node depth") +
          scale_y_log10() + 
          scale_fill_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) +
          theme_bw() + ggtitle('node depth threshold where accuracy >= 0.80') + theme(legend.position = "none")

# jscore_df4 %>% 
#           ggplot(aes(x=factor(mr), y=mean_node_height_at_50, fill = gaia_target)) +
#           #geom_jitter(width = 0.1, alpha = 0.1, size = 0.1) + 
#           geom_boxplot(alpha = 0.4) + 
#           facet_wrap(~pm, nrow = 1) + 
#           labs(x="migration strength", y="node depth") +
#           scale_y_log10() + 
#           scale_fill_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) +
#           theme_classic() + ggtitle('node depth threshold where accuracy >= 0.50') + theme(legend.position = "none")


```


```{r, fig.width = 7, fig.height = 5}

## boxplots where each point describes one run and each box describes an mr 

jscore_df4 %>% 
          ggplot(aes(x=factor(mr), y=mean_node_height_at_80, fill = gaia_target)) +
          #geom_jitter(width = 0.1, alpha = 0.1, size = 0.1) + 
          geom_boxplot(alpha = 0.4) + 
          #facet_wrap(~pm, nrow = 1) + 
          labs(x="migration strength", y="node depth") +
          scale_y_log10(breaks = c(1, 10, 100, 1000, 10000, 100000)) + 
          scale_fill_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) +
          theme_bw() + 
          ggtitle('node depth threshold where accuracy >= 0.80') + theme(legend.position = "none")

# 
# jscore_df4 %>% 
#           ggplot(aes(x=factor(mr), y=mean_node_height_at_50, fill = gaia_target)) +
#           #geom_jitter(width = 0.1, alpha = 0.1, size = 0.1) + 
#           geom_boxplot(alpha = 0.4) + 
#           #facet_wrap(~pm, nrow = 1) + 
#           labs(x="migration strength", y="node depth") +
#           scale_y_log10(breaks = c(1, 10, 100, 1000, 10000, 100000)) + 
#           scale_fill_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) +
#           theme_classic() + 
#         ggtitle('node depth threshold where accuracy >= 0.50') + theme(legend.position = "none")



```





```{r}

#heatmap 

# jscore_df4 %>% group_by(pm, mr, gaia_target) %>% summarise(threshold80 = mean(mean_node_height_at_80, na.rm = TRUE), .groups = 'drop') %>% 
#           ggplot(aes(x=factor(mr), y=factor(pm), fill = threshold80)) +
#           geom_tile() + coord_fixed() + 
#           scale_fill_gradient(low = "blue", high = "red") + facet_wrap(~gaia_target) +
#           labs(x="migration strength", y="p/m") +
#           #scale_y_log10() + 
#           #scale_fill_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) +
#           theme_classic() + ggtitle('node depth threshold where accuracy >= 0.80') #+ theme(legend.position = "none")

```



