re cecile comments to plot_logistic_gaia.Rmd 

# viz for gaia accuracy testing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load libraries
```{r}
library(gaia)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(stringr)
library(purrr)
library(scales)
library(gridExtra)
library(rsample)
```

## set work dir
```{r}
setwd("/home/nahmed/git/bacterial-args/migration/sweep_height_long/")
```

## ggplot theme
```{r}
## ggplot theme

my_theme <- function() {
    theme_classic() +
    theme( 
        plot.title = element_text(color="black", size=12, face="bold"),
        axis.title.x = element_text(color="black", size=20, face="bold", vjust=2),
        axis.title.y = element_text(color="black", size=20, face="bold", vjust=2),
        axis.text=element_text(size=24),
        )
}

options(repr.plot.width=14, repr.plot.height=8)
```

## read in scores df
```{r}
scores_eq_pairs <- read.csv("scores_eq_pairs_003.csv")
scores_eq_pairs <- scores_eq_pairs %>% mutate(node_height_sim = node_time) %>% mutate(gaia_target = recode(gaia_target, "inf" = "inferred", "sim" = "simulated"))
scores_eq_pairs$gaia_target <- factor(scores_eq_pairs$gaia_target, levels = c("simulated", "inferred"))
scores_eq_pairs$mr <- factor(scores_eq_pairs$mr, 
                                      levels=c("1e-05","0.001258" ,"0.002505", "0.003753", "0.005"), 
                                      labels = c("1e-5", "1.25e-3", "2.5e-3", "3.75e-3", "5e-3" ) )
scores_eq_pairs$pm <- factor(scores_eq_pairs$pm, 
                                      levels=c(.00100,0.07575,0.15050, 0.22530, 0.30000), 
                                      labels = c(".001", ".076", ".151", ".226", "0.3" ) )

#write.csv(scores_eq_pairs, "scores_eq_pairs_004.csv")

```


## extract root height by finding max node_height per-bin 
later: crude scoring of runs according to root labeling accuracy 
```{r}
roots <- scores_eq_pairs %>% group_by(bin, mr, mu, pm, seed, gaia_target) %>% summarise(root_height = max(node_height_sim, na.rm = TRUE)) %>% select(-bin)

roots <- roots %>% left_join(scores_eq_pairs, by = c('bin', 'mr', 'mu', 'pm', 'seed', 'gaia_target', "root_height" = "node_height_sim"))

```


# plot accuracy x nodeheight across pm and mr 
```{r, fig.width = 12, fig.height = 8}
scores_eq_pairs %>% group_by(pm, mr, node_height_sim, gaia_target) %>% summarise(score = mean(score, na.rm = TRUE)) %>% ungroup() %>% ggplot(aes(x=node_height_sim, y=score, color=gaia_target)) + 
                    geom_point(alpha = 0.75) + 
                    facet_wrap(pm~mr) + 
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    theme_classic() + theme(legend.position = "none")

#ggsave("accuracy_nodeheight_5x5.png")
```


## select specific p/m mu mr run(s)
```{r}

scores_eq_pairs %>% filter((pm == '.001') & (mr == '1e-5')) %>% select(seed) %>% unique() 

```

## EX. simple [logistic regression](https://koalaverse.github.io/homlr/notebooks/05-logistic-regression.nb.html)
```{r}

attrition <- modeldata::attrition
df <- attrition %>% mutate_if(is.ordered, factor, ordered = FALSE)

churn_split <- initial_split(df, prop = .7, strata = "Attrition")
churn_train <- training(churn_split)

model1 <- glm(Attrition ~ MonthlyIncome, family = "binomial", data = churn_train)
#model2 <- glm(Attrition ~ OverTime, family = "binomial", data = churn_train)

churn_train2 <- churn_train %>% mutate(prob = ifelse(Attrition == "Yes", 1, 0))
churn_train2 <- broom::augment(model1, churn_train2) %>% mutate(.fitted = exp(.fitted))

ggplot(churn_train2, aes(MonthlyIncome, prob)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  ggtitle("Predicted probabilities for model1") +
  xlab("Monthly Income") +
  ylab("Probability of Attrition")


tidy(model1)
as.data.frame(exp(coef(model1))) # odds ratios
as.data.frame(confint(model1)) # confidence intervals
tidy(predict(object = model1, type = "response"))

```
## EX. cécile
```{r}

x <- 1:10
y <- c(rep(0, 5), rep(1, 5))
df_data <- tibble(x = x, y = y)

log_reg <- glm(data = df_data, formula = y ~ x, family = 'binomial')
df_pred <- tibble(x = seq(0, 12, 0.001))

df_pred$y_pred <- predict(object = log_reg, newdata = df_pred, type = "response")

## supply a grid of node_depths, predict accuracies for them 
## find first node depth for which accuracy drops below 80% -- get uncertainty, find the node depth for which lower bound of accuracy dips below 80
## get half life from me eventually .... 
## if problem is duplication of points, think of strategies for getting rid of redunancy? idk. 
## just test on a sim whether deduplication solves weird curve issue. 
## testing clade level mrca scoring rather than pairwise -- choose one network 

df_data2 <- broom::augment(log_reg, df_data) %>% mutate(.fitted = exp(.fitted))


df_data %>% 
  ggplot(aes(x = x)) +
  geom_point(aes(y = y)) + geom_line(data = df_pred, aes(y = y_pred))

ggplot(df_data2, aes(x=x, y=y)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  ggtitle("Predicted probabilities for glm") +
  xlab("x") +
  ylab("y")



```
### apply to single run for particular pm and mr:
```{r}

test <- scores_eq_pairs %>% filter((pm == '.001') & (mr == '1e-5')  & (seed == 140) & (gaia_target == 'simulated'))

test %>% 
        ggplot(aes(x=node_height_sim, y=score, color=gaia_target)) + 
                    geom_point(alpha = 0.05, color = 'black') + 
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                   # scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    theme_classic() 

```

```{r}

test %>% 
  ggplot(aes(x = node_height_sim, fill = as.factor(score))) + geom_histogram(position = 'stack', binwidth = 0.1) + scale_x_log10() + theme_classic() + 
            xlab("node depth") + ylab('count') + scale_fill_manual(values = c('grey', 'black'))

```

```{r}
run <- test
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme() 

tidy(model1) # high standard error, model does not converge
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```
same data using cecile's example
```{r}


## find first node depth for which accuracy drops below 80% -- get uncertainty, find the node depth for which lower bound of accuracy dips below 80
## get half life from me eventually .... 
## if problem is duplication of points, think of strategies for getting rid of redunancy? idk. 
## just test on a sim whether deduplication solves weird curve issue. 
## testing clade level mrca scoring rather than pairwise -- choose one network 


# df = 'run' 
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run_pred <- tibble(node_height_sim = seq(0.5, 2e5), 50) ## supply a grid of node_depths, predict accuracies for them 
run_pred$score_pred <- predict(object = model1, newdata = run_pred, type = "response")
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))


run2 %>% 
  ggplot(aes(x = node_height_sim, y = score)) + 
  
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se = FALSE, linewidth = 2, color = 'lightblue') + 
  geom_point() + 
  geom_line(data = run_pred, aes(y = score_pred), linewidth = 1) + 
  scale_x_log10() + geom_hline(yintercept = .80, color = "grey") + 
  geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
  my_theme() + xlab('node depth') + ylab('accuracy') 
  
```
predicted fit looks great. 


# investigating weird run from before, where geom_smooth() fit seems inaccurate
prediction is a lot more confident about the larger node_height, possibly due to duplication.. weirdly low predictions for lower node_height
```{r}
runw <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated') & (seed == 45) )
model1 <- glm(score ~ node_height_sim, family = "binomial", data = runw)
runw_pred <- tibble(node_height_sim = seq(0.5, 1e4), 50) ## supply a grid of node_depths, predict accuracies for them 
runw_pred$score_pred <- predict(object = model1, newdata = runw_pred, type = "response")
runw2 <- broom::augment(model1, runw) %>% mutate(.fitted = exp(.fitted))

runw2 %>% 
  ggplot(aes(x = node_height_sim, y = score)) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se = FALSE, linewidth = 2, color = 'lightblue') + 
  geom_point() + 
  geom_line(data = runw_pred, aes(y = score_pred), linewidth = 1) + 
  scale_x_log10() + geom_hline(yintercept = .80, color = "grey") + 
  geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
  my_theme() + xlab('node depth') + ylab('accuracy') 



runw2 %>% 
  ggplot(aes(x = node_height_sim, fill = as.factor(score))) + geom_histogram(position = 'stack', binwidth = 0.1) + scale_x_log10() + theme_classic() + 
            xlab("node depth") + ylab('count') + scale_fill_manual(values = c('grey', 'black')) 
#ggsave('original_hist.png')

```
only predicting within range of original dataset
```{r}

runw <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated') & (seed == 45) )
model1 <- glm(score ~ node_height_sim, family = "binomial", data = runw)
runw_pred <- tibble(node_height_sim = seq(min(runw$node_height_sim), max(runw$node_height_sim), 0.01)) ## supply a grid of node_depths, predict accuracies for them 
runw_pred$score_pred <- predict(object = model1, newdata = runw_pred, type = "response")
runw2 <- broom::augment(model1, runw) %>% mutate(.fitted = exp(.fitted)) 

runw2 %>% 
  ggplot(aes(x = node_height_sim, y = score)) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se = FALSE, linewidth = 2, color = 'lightblue') + 
  geom_point() + 
  geom_line(data = runw_pred, aes(y = score_pred), linewidth = 1) + 
  scale_x_log10() + geom_hline(yintercept = .80, color = "grey") + 
  geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
  my_theme() + xlab('node depth') + ylab('accuracy') + ggtitle('y~x') 

#ggsave('original.png')

```
# fit to log(node depth) - this matches whats going on with geom_smooth .. still bad
```{r}
runw <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated') & (seed == 45) )
model1 <- glm(score ~ log10(node_height_sim), family = "binomial", data = runw)
runw_pred <- tibble(node_height_sim = seq(min(runw$node_height_sim), max(runw$node_height_sim)), 50) ## supply a grid of node_depths, predict accuracies for them 
runw_pred$score_pred <- predict(object = model1, newdata = runw_pred, type = "response")
runw2 <- broom::augment(model1, runw) %>% mutate(.fitted = exp(.fitted))

runw2 %>% 
  ggplot(aes(x = node_height_sim, y = score)) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se = FALSE, linewidth = 2.2, color = 'lightblue') + 
  geom_point() + 
  geom_line(data = runw_pred, aes(y = score_pred), linewidth = 1) + 
  scale_x_log10() + geom_hline(yintercept = .80, color = "grey") + 
  geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
  my_theme() + xlab('node depth') + ylab('accuracy') + ggtitle('y~log(x)')

#ggsave('log.png')
```

# try deduplication to improve score fitting ... 
```{r}
runw <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated') & (seed == 45) )
runw <- runw %>% group_by(node, node_height_sim, score) %>% count(node) %>% arrange(node_height_sim) # deduplicate
model1 <- glm(score ~ node_height_sim, family = "binomial", data = runw)
runw_pred <- tibble(node_height_sim = seq(min(runw$node_height_sim), max(runw$node_height_sim)), 50) ## supply a grid of node_depths, predict accuracies for them 
runw_pred$score_pred <- predict(object = model1, newdata = runw_pred, type = "response")
runw2 <- broom::augment(model1, runw) %>% mutate(.fitted = exp(.fitted))

runw

runw2 %>% 
  ggplot(aes(x = node_height_sim, y = score)) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se = FALSE, linewidth = 2, color = 'lightblue') + 
  geom_point() + 
  geom_line(data = runw_pred, aes(y = score_pred), linewidth = 1) + 
  scale_x_log10() + geom_hline(yintercept = .80, color = "grey") + 
  geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
  my_theme() + xlab('node depth') + ylab('accuracy') + ggtitle('deduplicated scores, y~x')

#ggsave('dedup.png')

runw2 %>% 
  ggplot(aes(x = node_height_sim, fill = as.factor(score))) + geom_histogram(position = 'stack', binwidth = 0.1) + scale_x_log10() + theme_classic() + 
            xlab("node depth") + ylab('count') + scale_fill_manual(values = c('grey', 'black')) 

#ggsave('dedup_hist.png')

```
# deduplication + doing y~log(x) 
```{r}
runw <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated') & (seed == 45) )
runw <- runw %>% group_by(node, node_height_sim, score) %>% count(node) %>% arrange(node_height_sim) # deduplicate
model1 <- glm(score ~ log10(node_height_sim), family = "binomial", data = runw)
runw_pred <- tibble(node_height_sim = seq(min(runw$node_height_sim), max(runw$node_height_sim)), 50) ## supply a grid of node_depths, predict accuracies for them 
runw_pred$score_pred <- predict(object = model1, newdata = runw_pred, type = "response")
runw2 <- broom::augment(model1, runw) %>% mutate(.fitted = exp(.fitted))

runw

runw2 %>% 
  ggplot(aes(x = node_height_sim, y = score)) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se = FALSE, linewidth = 2, color = 'lightblue') + 
  geom_point() + 
  geom_line(data = runw_pred, aes(y = score_pred), linewidth = 1) + 
  scale_x_log10() + geom_hline(yintercept = .80, color = "grey") + 
  geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
  my_theme() + xlab('node depth') + ylab('accuracy') + ggtitle('deduplicated scores, y~log(x)')

#ggsave('dedup_log.png')

runw2 %>% 
  ggplot(aes(x = node_height_sim, fill = as.factor(score))) + geom_histogram(position = 'stack', binwidth = 0.1) + scale_x_log10() + theme_classic() + 
            xlab("node depth") + ylab('count') + scale_fill_manual(values = c('grey', 'black')) 
            


```
Deduplication + modeling on a log scale seems to work better. Deduplication makes sense when gaia_target = sim because each scored node corresponds to a true node. I am weary about deduplicating for inferred tree sequences. Filtering by clade-accuracy will help to reduce some duplication but not all of it. Not sure where to go from here. 



# function 
```{r}


# return df with cols 
# uncertainty lower and upper bounds ; predicted score, upper bound, lower bound, pm, mr, target  
# find first node depth for which accuracy drops below 80% -- get uncertainty, find the node depth for which lower bound of accuracy dips below 80

predict(model1, )
## function to return r2, intercept, slope .... predict df, params, etc. --  loop over every set of inputs 

log10(node_height_sim) # try fitting model on log scale node_height data


```

