# viz for gaia accuracy testing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load libraries
```{r}
library(gaia)
library(tidyr)
library(dplyr)
library(data.table)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(RColorBrewer)
library(viridis)
library(stringr)
library(purrr)
library(scales)
```

## set work dir
```{r}
setwd("/home/nahmed/git/bacterial-args/migration/sweep_height_long/")
```

## ggplot theme
```{r}
## ggplot theme

my_theme <- function() {
    theme_classic() +
    theme( 
        plot.title = element_text(color="black", size=12, face="bold"),
        axis.title.x = element_text(color="black", size=20, face="bold", vjust=2),
        axis.title.y = element_text(color="black", size=20, face="bold", vjust=2),
        axis.text=element_text(size=24),
        )
}

options(repr.plot.width=14, repr.plot.height=8)
```

## read in scores df
```{r}
scores_eq_pairs <- read.csv("scores_eq_pairs_003.csv")
scores_eq_pairs <- scores_eq_pairs %>% mutate(node_height_sim = node_time) %>% mutate(gaia_target = recode(gaia_target, "inf" = "inferred", "sim" = "simulated"))
scores_eq_pairs$gaia_target <- factor(scores_eq_pairs$gaia_target, levels = c("simulated", "inferred"))
scores_eq_pairs$mr <- factor(scores_eq_pairs$mr, 
                                      levels=c("1e-05","0.001258" ,"0.002505", "0.003753", "0.005"), 
                                      labels = c("1e-5", "1.25e-3", "2.5e-3", "3.75e-3", "5e-3" ) )
scores_eq_pairs$pm <- factor(scores_eq_pairs$pm, 
                                      levels=c(.00100,0.07575,0.15050, 0.22530, 0.30000), 
                                      labels = c(".001", ".076", ".151", ".226", "0.3" ) )

#write.csv(scores_eq_pairs, "scores_eq_pairs_004.csv")

```


## extract root height by finding max node_height per-bin 
later: crude scoring of runs according to root labeling accuracy 
```{r}
roots <- scores_eq_pairs %>% group_by(bin, mr, mu, pm, seed, gaia_target) %>% summarise(root_height = max(node_height_sim, na.rm = TRUE)) %>% select(-bin)

roots <- roots %>% left_join(scores_eq_pairs, by = c('bin', 'mr', 'mu', 'pm', 'seed', 'gaia_target', "root_height" = "node_height_sim"))

```


# plot accuracy x nodeheight across pm and mr 
```{r, fig.width = 12, fig.height = 8}
scores_eq_pairs %>% group_by(pm, mr, node_height_sim, gaia_target) %>% summarise(score = mean(score, na.rm = TRUE)) %>% ungroup() %>% ggplot(aes(x=node_height_sim, y=score, color=gaia_target)) + 
                    geom_point(alpha = 0.75) + 
                    facet_wrap(pm~mr) + 
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    theme_classic() + theme(legend.position = "none")

ggsave("accuracy_nodeheight_5x5.png")
```


## select specific p/m mu mr run(s)
```{r}

scores_eq_pairs %>% filter((pm == '.001') & (mr == '1e-5')) %>% select(seed) %>% unique() 

```

## EX. simple [logistic regression](https://koalaverse.github.io/homlr/notebooks/05-logistic-regression.nb.html)
```{r}

attrition <- modeldata::attrition
df <- attrition %>% mutate_if(is.ordered, factor, ordered = FALSE)

churn_split <- initial_split(df, prop = .7, strata = "Attrition")
churn_train <- training(churn_split)

#model1 <- glm(Attrition ~ MonthlyIncome, family = "binomial", data = churn_train)
model2 <- glm(Attrition ~ OverTime, family = "binomial", data = churn_train)

churn_train2 <- churn_train %>% mutate(prob = ifelse(Attrition == "Yes", 1, 0))
churn_train2 <- broom::augment(model2, churn_train2) %>% mutate(.fitted = exp(.fitted))

ggplot(churn_train2, aes(MonthlyIncome, prob)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  ggtitle("Predicted probabilities for model1") +
  xlab("Monthly Income") +
  ylab("Probability of Attrition")


tidy(model2)
as.data.frame(exp(coef(model2)))
as.data.frame(confint(model2))

```

# 01
### apply to single run for particular pm and mr:
```{r}

scores_eq_pairs %>% filter((pm == '.001') & (mr == '1e-5')  & (seed == 140) & (gaia_target == 'simulated')) %>% 
        ggplot(aes(x=node_height_sim, y=score, color=gaia_target)) + 
                    geom_point(alpha = 0.05, color = 'black') + 
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                   # scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    theme_classic() 

```
### lowest pm, lowest migration strength 
```{r}

run <- scores_eq_pairs %>% filter((pm == '.001') & (mr == '1e-5')  & (seed == 140) & (gaia_target == 'simulated'))
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme() 

tidy(model1) # ridiculously high standard error, model does not converge, etc. 
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```

### apply to all runs for particular pm and mr 
```{r}

run <- scores_eq_pairs %>% filter((pm == '.001') & (mr == '1e-5') & (gaia_target == 'simulated'))
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme()

tidy(model1)
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```
# 02
### slightly increasing p/m and migration strength 

#### single run 
converges fine and doesn't have ridiculous se -- *what's the difference between this and the prior single networks?* that predictor and response are basically perfectly correlated - after a certain node_height threshold, every score value is 0. 
```{r}

run <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated') & (seed == 45) )
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme()

tidy(model1)
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```

#### all runs
```{r}

run <- scores_eq_pairs %>% filter((pm == '.076') & (mr == '1.25e-3') & (gaia_target == 'simulated'))
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme()

tidy(model1)
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```
# 03
### slightly increasing p/m and migration strength 

#### single run 
converges fine and doesn't have ridiculous se -- *what's the difference between this and the prior single networks?* that predictor and response are basically perfectly correlated - after a certain node_height threshold, every score value is 0. 
```{r}

run <- scores_eq_pairs %>% filter((pm == '.151') & (mr == '2.5e-3') & (gaia_target == 'simulated') & (seed == 76) )
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme()

tidy(model1)
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```

#### all runs
```{r}

run <- scores_eq_pairs %>% filter((pm == '.151') & (mr == '2.5e-3') & (gaia_target == 'simulated'))
model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))

run2 %>% 
        ggplot(aes(x=node_height_sim, y=score)) + 
                    geom_point(alpha = 0.15) + 
                    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                    #facet_wrap(~mr) + 
                    labs(x="node_depth", y="accuracy", color="gaia target") + 
                    scale_x_log10() + 
                    geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                    #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                    my_theme()

tidy(model1)
as.data.frame(exp(coef(model1)))
as.data.frame(confint(model1))

```

# multi plot

```{r}
#library(gridExtra)

pms <- unique(scores_eq_pairs$pm)
mrs <- unique(scores_eq_pairs$mr)

plot_list <- list() 

#par(mfrow = c(3, 2))
for (i in 1:length(pms)) {
  
  run <- scores_eq_pairs %>% filter((pm == pms[i]) & (mr == mrs[i]) & (gaia_target == 'simulated'))
  model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
  run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))
  
  p <- run2 %>% 
          ggplot(aes(x=node_height_sim, y=score)) + 
                      geom_point(alpha = 0.15) + 
                      geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                      #facet_wrap(~mr) + 
                      labs(x="node_depth", y="accuracy", color="gaia target") + 
                      scale_x_log10() + 
                      geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                      #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                      theme_classic() + ggtitle(paste0("p/m: ", pms[i], ", mr: ", mrs[i]))
  
  plot_list[[i]] <- p
}

do.call(grid.arrange, c(plot_list, nrow = 2, ncol = 3  ))

```
```{r}

plot_list_all <- list() 
count <- 1

for (i in 1:length(pms)) {
  for (j in 1:length(mrs)) {
  
    run <- scores_eq_pairs %>% filter((pm == pms[i]) & (mr == mrs[j]) & (gaia_target == 'simulated')) 
    seed <- sample(unique(run$seed), 1)
    run <- run %>% filter(seed == seed)
    model1 <- glm(score ~ node_height_sim, family = "binomial", data = run)
    run2 <- broom::augment(model1, run) %>% mutate(.fitted = exp(.fitted))
    
    
    p <- run2 %>% ggplot(aes(x=node_height_sim, y=score)) + 
                        geom_point(alpha = 0.15) + 
                        geom_smooth(method = "glm", method.args = list(family = "binomial")) +
                        #facet_wrap(~mr) + 
                        labs(x="node_depth", y="accuracy", color="gaia target") + 
                        scale_x_log10() + 
                        geom_hline(yintercept = .80, color = "grey") + geom_hline(yintercept = .50, linetype = "dashed", color = "grey") +
                        #scale_color_manual(values = c("simulated" = "#F2AD00", "inferred" = "#00A08A")) + 
                        theme_classic() + 
                        ggtitle(paste0("p/m: ", pms[i], ", mr: ", mrs[j], ", seed: ", seed)) + 
                        theme(plot.title = element_text(size = 8),
                              plot.subtitle = element_text(size = 8), 
                              axis.title = element_text(size = 8),
                              axis.text = element_text(size = 6))
    
    plot_list_all[[count]] <- p
    count <- count + 1
    
  }
}

do.call(grid.arrange, c(plot_list_all, nrow = 5, ncol = 5))
#ggsave('logistic_all.png', width = 12, height = 12)

```

